Data Specialization for a Staged Lambda Calculus

### 2015-01-22 paper/example.tex
small changes to section 2

We can formalize this intuition by writing quickselect in a \emph{staged}
language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
lambda calculus. This allows us to directly express the idea that the argument
\texttt{l} is known at the \emph{first} stage of the computation, while the
argument \texttt{k} is only known at the \emph{second}, and all
purely-stage-\bbone\ computation occurs before any stage \bbtwo\ computation.

We define a staged version of quickselect in \ref{fig:qs-staged}, writing stage
\bbone\ code in red, and stage \bbtwo\ code in blue. Whereas \texttt{qSelect}
sends a \textrm{list} and \rmint\ to an \rmint, \texttt{qsStaged} is more
precise---it is a stage \bbone\ function which takes a \textrm{list} and a
$\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

@@

The process of adding staging annotations ($\fut$ types, $\next$, $\prev$, and
$\pause$) to unstaged code has been the subject of extensive research under the
name of \emph{binding time analysis}. In this paper, we assume that these
annotations have been provided by the programmer (or perhaps a binding time
analysis tool), and do not consider the problem of generating such annotations.
There are many ways to annotate any program, including \texttt{qSelect}; we
chose annotations which maximize the work performed at stage \bbone.

@@

An astute programmer, having noticed that \texttt{quickselect} can be staged in
this fashion, might try to split it into a pair of functions, one which performs
all the work depending only on \texttt{l} (stage \bbone), and one which uses
\texttt{k} and the result of that work on \texttt{l} to compute the element with
rank \texttt{k} in \texttt{l}.

@@

@@

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
produces a pair of terms $e_1,e_2$ containing respectively all the stage \bbone\
and stage \bbtwo\ content of $e$, such that $e_2~e_1$ is equivalent to $e$.
(Such a splitting is always possible because $e$ is well-typed in \lang.) In the
case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
described above.

>>>>

We can formalize this intuition by writing quickselect in a \emph{staged}
language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
lambda calculus. This allows us to directly express the idea that the argument
\texttt{l} is known at the \emph{first} stage of the computation, while the
argument \texttt{k} is only known at the \emph{second}, and all
first-stage computation occurs before any second-stage computation.

We define a staged version of quickselect in \ref{fig:qs-staged}, writing
first-stage computations in red, and second-stage computations in blue. While
\texttt{qSelect} sends a \textrm{list} and \rmint\ to an \rmint,
\texttt{qsStaged} has a more precise type---it is a first-stage function which
takes a \textrm{list} and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and
returns a $\fut\rmint$.

\crem{Next three paragraphs are waiting on exact code for \ref{fig:qs-staged}.
Use three colors?}

@@

The process of adding staging annotations ($\fut$ types, $\next$, $\prev$, and
$\pause$) to unstaged code has been the subject of extensive research under the
name of \emph{binding time analysis}. In this paper, we assume that these
annotations have been provided by the programmer (or perhaps a binding time
analysis tool), and do not consider the problem of generating such annotations.
There are many ways to annotate any program, including \texttt{qSelect}; we
chose annotations which maximize the work performed in the first stage.

@@

An astute programmer, having noticed that \texttt{quickselect} can be staged in
this fashion, might try to split it into a pair of functions, one which performs
all the work depending only on \texttt{l} (the first stage), and one which uses
\texttt{k} and the result of that work on \texttt{l} to compute the element with
rank \texttt{k} in \texttt{l}.

@@

\crem{Use $\bbonep$ to explain the above.}

@@

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
produces a pair of terms $e_1,e_2$ containing respectively all the first- and
second-stage content of $e$, such that $e_2~e_1$ is equivalent to $e$.
(Such a splitting is always possible because $e$ is well-typed in \lang.) In the
case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
described above.
